---
title: 'Supervised Learning Project: KNN vs Logistic Regression'
author: "Sami"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

# Introduction

This appendix presents the R implementation used to preprocess the data, train the models, and evaluate their performance.

------------------------------------------------------------------------

# 1. Load Required Packages

The following packages are required for KNN classification and ROC curve analysis.

```{r}
library(class)
library(pROC)

```

# 2. Load and Inspect Dataset

We load the dataset and assign column names for better readability.

```{r cars}
setwd("C:/Users/SYB HP/Desktop/statistical_learn_project/supervised")

data <- read.csv("adult.csv", header = FALSE, strip.white = TRUE)

colnames(data) <- c("age", "workclass", "fnlwgt", "education",
                    "education.num", "marital.status", "occupation",
                    "relationship", "race", "sex",
                    "capital.gain", "capital.loss",
                    "hours.per.week", "native.country", "income")

head(data)
str(data)
```

# 3. Data Preparation

Only numerical variables are selected to simplify the analysis. The target variable is converted into binary format.

```{r pressure, echo=FALSE}
data <- data[, c("age", "education.num", "hours.per.week",
                 "capital.gain", "capital.loss", "income")]

data$income <- ifelse(data$income == ">50K", 1, 0)
head(data)
```
```{r}
# EDA Income distribution

library(ggplot2)

ggplot(data, aes(x = income, fill = income)) +
  geom_bar() +
  labs(title = "Income Distribution",
       x = "Income Category",
       y = "Count")
```
```{r}

# EDA Correlation matrix

library(corrplot)

numeric_data <- data[, c("age",
                         "education.num",
                         "hours.per.week",
                         "capital.gain",
                         "capital.loss")]

cor_matrix <- cor(numeric_data)

corrplot(cor_matrix,
         method = "color",
         addCoef.col = "black")
```

# 4. Train-Test Split

The dataset is split into training (70%) and testing (30%) sets.

```{r}
set.seed(123)
index <- sample(1:nrow(data), 0.7*nrow(data))

train <- data[index, ]
test <- data[-index, ]
```

# 5. Data Normalization

Min-max scaling is applied to ensure all variables are on the same scale. This step is crucial for KNN.

```{r}
set.seed(123)

scale_data <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

train_scaled <- train
test_scaled <- test

train_scaled[,1:5] <- scale_data(train[,1:5])
test_scaled[,1:5] <- scale_data(test[,1:5])
```

# 6. Logistic Regression Model

We train a logistic regression model and compute prediction accuracy.

```{r}
set.seed(123)

model_log <- glm(income ~ ., data=train, family=binomial)

prob_log <- predict(model_log, test, type="response")

pred_log <- ifelse(prob_log > 0.5, 1, 0)

accuracy_log <- mean(pred_log == test$income)
accuracy_log
```

# 7. KNN Model

We apply the KNN algorithm using normalized data.

```{r}
set.seed(123)

train_x <- train_scaled[,1:5]
test_x <- test_scaled[,1:5]
train_y <- train$income

pred_knn <- knn(train_x, test_x, train_y, k=5)

accuracy_knn <- mean(pred_knn == test$income)
accuracy_knn
```

```{r}
```

# 8. Evaluation Metrics (Logistic Regression)

We compute confusion matrix, precision, recall, and F1-score.

```{r}
set.seed(123)

cm_log <- table(Predicted = pred_log, Actual = test$income)


TN <- cm_log[1,1]
FN <- cm_log[1,2]
FP <- cm_log[2,1]
TP <- cm_log[2,2]

precision_log <- TP / (TP + FP)
recall_log <- TP / (TP + FN)
f1_log <- 2 * (precision_log * recall_log) / (precision_log + recall_log)
"Confusion matrix Log:"
print(cm_log)
print(paste("Precision Logistic:", precision_log))
print(paste("Recall Logistic:", recall_log))
print(paste("F1 Logistic:", f1_log))

```

# 9. Evaluation Metrics (KNN)

Same evaluation is performed for KNN.

```{r}
set.seed(123)

cm_knn <- table(Predicted = pred_knn, Actual = test$income)

TN <- cm_knn[1,1]
FN <- cm_knn[1,2]
FP <- cm_knn[2,1]
TP <- cm_knn[2,2]

precision_knn <- TP / (TP + FP)
recall_knn <- TP / (TP + FN)
f1_knn <- 2 * (precision_knn * recall_knn) / (precision_knn + recall_knn)

"Confusion matrix knn:"
cm_knn

print(paste("Precision KNN:", precision_knn))
print(paste("Recall KNN:", recall_knn))
print(paste("F1 KNN:", f1_knn))
```

# 10. K Optimization

We test different values of K to find the optimal configuration.

```{r}
set.seed(123)

for (k in 1:10) {
  pred <- knn(train_x, test_x, train_y, k=k)
  print(paste("k =", k, "accuracy =", mean(pred == test$income)))

}
```

# 11. Accuracy vs K Plot

This plot shows how accuracy varies with K.

```{r}
set.seed(123)

k_values <- 1:10
acc <- c()

for (k in k_values) {
  pred <- knn(train_x, test_x, train_y, k=k)
  acc <- c(acc, mean(pred == test$income))
}

plot(k_values, acc, type="b",
     xlab="K value",
     ylab="Accuracy",
     main="KNN Accuracy vs K")

best_k <- k_values[which.max(acc)]

print(paste("Best k =", best_k))
```

# 12. ROC Curve and AUC

We evaluate model performance using ROC curves and AUC.

##Logistic Regression##

```{r}
set.seed(123)

roc_log <- roc(test$income, prob_log)

plot(roc_log, col="blue", main="ROC Curve Comparison")

auc(roc_log)
```

KNN

```{r}
set.seed(123)


# ROC Logistic (d'abord)
roc_log <- roc(test$income, prob_log)
plot(roc_log, col="blue", main="ROC Curve Comparison")

# ROC KNN (ensuite)
pred_knn_prob <- knn(train_x, test_x, train_y, k=best_k, prob=TRUE)

prob_knn <- ifelse(pred_knn_prob == 1,
                   attr(pred_knn_prob, "prob"),
                   1 - attr(pred_knn_prob, "prob"))

roc_knn <- roc(test$income, prob_knn)

plot(roc_knn, col="red", add=TRUE)

# AUC
auc(roc_log)
auc(roc_knn)

# Legend
legend("bottomright",
       legend=c("Logistic", "KNN"),
       col=c("blue", "red"),
       lwd=2)
```
