---
title: "Unsupervised Learning Project: Customer Segmentation using Clustering"
author: "Sami"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

# Introduction

This appendix presents the R implementation used to preprocess the data, perform clustering, and analyze customer segments using unsupervised learning techniques.

The objective of this project is to identify natural groupings of customers based on their purchasing behavior without using any predefined target variable.

------------------------------------------------------------------------

# 1. Load Required Packages

The following packages are required for data manipulation, visualization, and clustering algorithms.

```{r setup, include=FALSE}

library(tidyverse)
library(factoextra)
library(cluster)
library(readxl)

```

# 2. Data Loading and Exploration

In this section, the dataset is loaded and explored to understand its structure and main characteristics.

The dataset contains transactional data from an online retail store, including information about invoices, products, quantities, prices, and customers.

Understanding the dataset is an important step before performing any preprocessing or clustering.

```{r cars}

setwd("C:/Users/SYB HP/Desktop/statistical_learn_project/unsupervised")

data <- read_excel("OnlineRetail.xlsx")

# Aperçu
head(data)
str(data)
summary(data)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
# delete missing values
data <- na.omit(data)

# Remove canceled transactions (Invoice start with C)
data <- data[!grepl("^C", data$InvoiceNo), ]

# delete negative quantities
data <- data[data$Quantity > 0, ]

# Verification
summary(data)
```
```{r}
# ================================
# BLOCK 4 — Feature Engineering
# ================================

# Create TotalPrice
data$TotalPrice <- data$Quantity * data$UnitPrice

#  date Conversion
data$InvoiceDate <- as.POSIXct(data$InvoiceDate,
                               format="%m/%d/%Y %H:%M")

# Verification
head(data)
```

```{r pressure, echo=FALSE}

```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 5 — Customer Aggregation
# ================================

data_customer <- data %>%
  group_by(CustomerID) %>%
  summarise(
    total_spent = sum(TotalPrice),
    frequency = n(),
    avg_spent = mean(TotalPrice)
  )

# Remove rows with missing CustomerID
data_customer <- na.omit(data_customer)

head(data_customer)
```
```{r}
# ================================
# EDA — Total Spending Distribution
# ================================
ggplot(data_customer,
       aes(x = total_spent)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Total Spending",
       x = "Total Spending",
       y = "Frequency")

```


```{r}

# ================================
# EDA — Purchase Frequency
# ================================

ggplot(data_customer,
       aes(x = frequency)) +
  geom_histogram(bins = 30) +
  labs(title = "Purchase Frequency Distribution",
       x = "Frequency",
       y = "Count")
```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 6 — Normalization
# ================================

data_scaled <- scale(data_customer[, -1])

# Verification
head(data_scaled)
```

```{r pressure, echo=FALSE}

# ================================
# BLOCK 7 — Elbow Method
# ================================

wss <- numeric(10)

for (k in 1:10) {
  kmeans_model <- kmeans(data_scaled, centers = k, nstart = 25)
  wss[k] <- kmeans_model$tot.withinss
}

plot(1:10, wss, type="b",
     xlab="Number of Clusters",
     ylab="WSS",
     main="Elbow Method")
```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 8 — KMeans
# ================================

set.seed(123)

kmeans_model <- kmeans(data_scaled, centers = 3, nstart = 25)

# Ajouter cluster
data_customer$cluster <- kmeans_model$cluster
head(data_customer)
```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 9 — KMeans Visualization
# ================================

ggplot(data_customer, aes(x=total_spent, y=frequency,
                          color=factor(cluster))) +
  geom_point() +
  labs(title="KMeans Clustering",
       color="Cluster")
```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 10 — Hierarchical
# ================================

dist_matrix <- dist(data_scaled)

hc_model <- hclust(dist_matrix, method="ward.D2")
plot(hc_model, main="Dendrogram")
rect.hclust(hc_model, k = 7, border = "red")
```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 11 — Cut Tree
# ================================

hc_clusters <- cutree(hc_model, k=3)

data_customer$hc_cluster <- hc_clusters

head(data_customer)
```

```{r pressure, echo=FALSE}
# ================================
# BLOCK 12 — Visualization HC
# ================================

ggplot(data_customer,
       aes(x=total_spent, y=frequency,
           color=factor(hc_cluster))) +
  geom_point() +
  labs(title="Hierarchical Clustering",
       color="Cluster")
```

```{r pressure, echo=FALSE}


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
